---
title: 'Australia''s Environment Report: Biodiversity'
author: "Shoshana Rapley"
date: "2025-02-05"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load packages
pacman::p_load(galah, janitor, ozmaps, paletteer, readxl, sf, showtext, smoothr, terra, tidyterra, tidyverse)
```

# Background

Comprehensive data on changes in biodiversity, species distribution and abundance are either impossible or extremely expensive to collect. Here we summarise national, annual metrics for biodiversity that help us understand how biodiversity is faring.

# EPBC Lists

The Australian Government (DCCEEW) maintains lists of species and ecological communities at risk of extinction. We report on annual changes to the list and cumulative outcomes.

EPBC lists are maintained online on the SPRAT database. I applied the following data formatting steps in the excel file "epbc_listings_to2023.xlsx" (each of the following as tabs) prior to import into R:

* Flora and Fauna list (last updated 29/1/24): I copied the [threatened flora](https://www.environment.gov.au/cgi-bin/sprat/public/publicthreatenedlist.pl?wanted=flora ) and [threatened fauna](https://www.environment.gov.au/cgi-bin/sprat/public/publicthreatenedlist.pl?wanted=fauna) lists from the SPRAT webpage. I manually formatted these tables into long format. I used the currently accepted scientific name when two were listed and removed the naming authority.

* Changes to the list (last updated 29/1/24): I copied the [changes to the EPBC list](https://www.environment.gov.au/cgi-tmp/publiclistchanges.309265b1b3529d5bf13e.html ) from the SPRAT webpage. I used the 'text to columns' function in excel used to split common and binomial names (and removed naming authority). 

* Removals from the list (last updated 1/2/22): I copied the [removals from the EPBC list](https://www.environment.gov.au/cgi-tmp/publiclistchanges.407465b1b51cc727df3e.html) from the SPRAT webpage (page scrape done 1/2/22 and all subsequent additions manually added as new rows). I read all listing advice to determine whether listings were "true" removals i.e. an actual increase of the threatened taxa, rather than taxonomic change or new information. 

## Initial data wrangling

**Note! This section does not need to be done again. For 2024 onward I can append changes to the list to the epbc_processed_2023.csv file - see next section.**

Here I combine the current flora and fauna lists with changes to the list and removals from the list. 

```{r}
# Make table for initial listings of later removed species
initial <- read_excel("epbc_listings_to2023.xlsx", sheet = "Removals") %>%
  clean_names() %>%
  subset(grepl("Increase", reason) | grepl("Relisted", reason)) %>%
  select(c("group","species","name","previous_status","year_first_listed")) %>%
  rename(status = previous_status) %>%
  rename(effective = year_first_listed) %>%
  # add an initial listing date
  mutate(effective = gsub("2000",as_date("16/07/2000", format = "%d/%m/%Y"), effective))

# Make table for initial listings of later transferred species
transfers <- read_excel("epbc_listings_to2023.xlsx", sheet = "Changes") %>%
  clean_names() %>%
  filter(str_detect(status, "Transfer")) %>%
  mutate(status = word(status, start=3L)) %>%
  mutate(status = ifelse(str_detect(status, "Critically"),
                         "Critically Endangered", status)) %>%
  mutate(status = ifelse(str_detect(status, "Conservation"),
                         "Conservation Dependent", status)) %>%
  mutate(effective = as_date("16/07/2000", format = "%d/%m/%Y"))

# Join all data
epbc <- read_excel("epbc_listings_to2023.xlsx", sheet = "Fauna") %>%
  rbind(read_excel("epbc_listings_to2023.xlsx", sheet = "Flora")) %>%
  rbind(read_excel("epbc_listings_to2023.xlsx", sheet = "Changes")) %>%
  rbind(read_excel("epbc_listings_to2023.xlsx", sheet = "Delist")) %>%
  clean_names() %>%
  rbind(transfers) %>%
  rbind(initial) %>%
  arrange(species, status) %>%
  mutate(effective = as_date(effective)) %>%
  # standardise nomenclature and syntax, remove whitespace
  mutate(species = str_trim(species,side = c("both")))

# Lost some group data due to non standing naming so retrieving it here
group <- select(epbc, c(group, species)) %>% arrange(group) %>%
  filter(!duplicated(species)) %>% arrange(species)
  
epbc <- right_join(group, select(epbc, -(group)), by= "species") %>%
# Remove duplicates where transferred
  mutate(order = ifelse(str_detect(status, "Transfer"),1,2)) %>%
  arrange(order) %>%
  filter(!duplicated(cbind(species, effective))) %>%
  arrange(species) %>%
# Remove false initial listings
  arrange(desc(effective)) %>%
  filter(!duplicated(cbind(species, status))) %>%
  arrange(species) %>%
# Create column for year listed
  mutate(year = year(effective)) %>%
# Create column for listing type 
  mutate(listing = ifelse(grepl("to Extinct",status),"Transfer extinct",
                   ifelse(grepl("Transfer from Extinct",status),"Rediscovered",
                   ifelse(grepl("Extinct",status),"Extinct",
                   ifelse(grepl("Transfer",status),"Transfer",
                   ifelse(grepl("Delisted",status),"Delisted", "Listed")))))) %>%
  arrange(status) %>%
  subset(select = -c(order))

# Export 
write.csv(epbc, "epbc_temp_2023.csv", row.names = FALSE)
```

Manual tasks in exported "epbc_temp_2023.csv" file:

1) Check correct listings of the 2 species that were delisted and relisted:
  a) *Macroderma gigas*	Ghost Bat (true de and re listing) 
  Mammal	*Macroderma gigas*	Vulnerable	Ghost Bat	16/07/2000	2000	Listed
  b) *Amytornis textilis myall*	Western Grasswren (true de and re listing) 
  Bird	*Amytornis textilis myall*	Vulnerable	Western Grasswren 	16/07/2000	2000	Listed

2) Check listing of CI Pipistrelle as there seems to be a timeline issue in the official data; should go endangered 2001 -> uplist CI 2006 -> uplist extinct 2021

3) Manually check for duplicated common names (using conditional formatting) to find taxonomic synonyms to combine; usually occurs for species that were transferred but originally listed after 2000. Usually the initial listing has the outdated name (and therefore the group hasn't been brought across for the new name either).

4) Check duplicated listings in species where naming isn't consistent on the listings:
  a) *Genoplesium vernale* AKA *Corunastylis vernalis*
  b) *Galaxias truttaceus* (Western Australian population)
  c) *Phascolarctos cinereus* (combined populations of Qld, NSW and the ACT)

5) One species is missing from the current flora list but has no record of being delisted or removed from the list: *Macrozamia platyrhachis*, listed endangered 2000. [This seems to be fixed in 2024]

Next, I re-arrange the data so that all species are listed in every calendar year that they occur. This enables us to do an annual summary. 

```{r}
# read in data
epbc <- read_csv("epbc_temp_2023.csv") %>%
  mutate(status2 = gsub("Transfer from ..* to ", "", status),
         year = as.numeric(year)) %>%
  subset(!year==2024)

# add column for number of times listed
epbc_count <- epbc %>%
  group_by(species) %>% 
  summarise(n = n()) 
epbc <- left_join(epbc, epbc_count)

# add years to 2023 for single listed species
sp <- unique(epbc$species)
epbc1 <- subset(epbc, n==1)
epbc1_full <- data.frame()

for(i in 1:length(sp)){
  temp <- filter(epbc1, species == sp[i])
  if(length(temp$year)==0) {
    next
  }
  yrs <- data.frame(species = sp[i],
                    year2 = temp$year:2023)
  out <- left_join(temp, yrs)
  epbc1_full <- rbind(epbc1_full, out)
}

# add years to listing change and to 2023 for species that changed status
epbc2 <- subset(epbc, n==2)
epbc2_dup <- distinct(epbc2, species, .keep_all = TRUE)
epbc2_full <- data.frame()

for(i in 1:length(sp)){
  temp <- filter(epbc2, species == sp[i])
  if(nrow(temp)==0) {
    next
  }
  status <- unique(temp$status2)
  listed <- unique(temp$year)
  change1 <- data.frame(species = sp[i],
                        status2 = status[1],
                       year2 = listed[1]:(listed[2]-1))
  change2 <- data.frame(species = sp[i],
                        status2 = status[2],
                        year2 = listed[2]:2023)
  changes <- rbind(change1, change2)
  line <- filter(epbc2_dup, species == sp[i]) %>%
    select(!status2)
  out <- left_join(line, changes, by = join_by(species))
  epbc2_full <- rbind(epbc2_full, out)
}

# add years to listing change and to 2023 for species that changed status twice
epbc3 <- subset(epbc, n==3)
epbc3_dup <- distinct(epbc3, species, .keep_all = TRUE)
epbc3_full <- data.frame()

for(i in 1:length(sp)){
  temp <- filter(epbc3, species == sp[i])
  if(nrow(temp)==0) {
    next
  }
  status <- temp$status2
  listed <- unique(temp$year)
  change1 <- data.frame(species = sp[i],
                        status2 = status[1],
                        year2 = listed[1]:(listed[2]-1))
  change2 <- data.frame(species = sp[i],
                        status2 = status[2],
                        year2 = listed[2]:(listed[3]-1))
  change3 <- data.frame(species = sp[i],
                        status2 = status[3],
                        year2 = listed[3]:2023)
  changes <- rbind(change1, change2, change3)
  line <- filter(epbc3_dup, species == sp[i]) %>%
    select(!status2)
  out <- left_join(line, changes, by = join_by(species))
  epbc3_full <- rbind(epbc3_full, out)
}

# rediscoveries
redisc <- unique(filter(epbc, listing=="Rediscovered")$species)

# combine all
epbc_all <- rbind(epbc1_full, epbc2_full, epbc3_full) %>%
  # remove years where delisted
  filter(!status2=="Delist") %>%
  select(c(group, species, name, status2, year2)) %>%
  rename(status = status2,
         year = year2) %>%
  # remove extinction status of rediscovered species
  mutate(temp = ifelse(species %in% redisc & status == "Extinct", 1,0)) %>%
  filter(temp==0) %>%
  select(!temp)

# write out
write.csv(epbc_all, "epbc_processed_2023.csv", row.names = FALSE)
```

Note to self: made an error in 2023. I had one species more than listed on the EPBC - which I narrowed down to being a vulnerable plant listing. I thought that Vappodes phalaenopsis ==	Dendrobium bigibbum but turns out they are indeed different listings. The duplicate I had instead was a name change for Helicteres sp. Glenluckie Creek	to Helicteres macrothrix - which looked like a new listing in 2006 instead of a name change, because of how it was listed on the changes to the EPBC list. I manually fixed this in "epbc_processed_2023.csv" on 14/2/2025.

## Annual data wrangling

For the 2023 report I did the initial data wrangling (see section above), so for all subsequent reports I can just append the changes in that calendar year.

I copied the annual updates from the [changes to the EPBC list](https://www.environment.gov.au/cgi-tmp/publiclistchanges.309265b1b3529d5bf13e.html) and the [removals from the EPBC list](https://www.environment.gov.au/cgi-tmp/publiclistchanges.268467a2e7a9bb580e3f.html) from the SPRAT webpage. I used the 'text to columns' function in excel used to split common and binomial names (and removed naming authority). I manually added the column "group".

Here we append the 2024 changes to the 2023 list. 

```{r}
# read in processed listings 2000-2023
epbc_all <- read_csv("data/epbc_processed_2023.csv")

# read in changes to the list in 2024
epbc_changes <- read_excel("data/epbc_changes_since2023_2024.xlsx", sheet = 1) %>%
  clean_names() %>%
  # drop effective date and add year
  select(!effective) %>%
  mutate(year = 2024) %>%
  # remove trailing whitespace 
  mutate(species = str_trim(species))

# read in delistings in 2024
epbc_remove <- read_excel("data/epbc_changes_since2023_2024.xlsx", sheet = 2) %>%
  clean_names()

remove <- epbc_remove$species

# bring across species from previous calendar year to current calendar year
epbc_prev <- epbc_all %>%
  filter(year == 2023) %>%
  mutate(year = 2024) %>%
  # remove species changed in 2024 (these added next step)
  filter(!species %in% epbc_changes$species)

# combine full listings with changes to the list
epbc_2024 <- rbind(epbc_changes, epbc_prev) %>%
  # where duplicates, keep the transfer listing
  distinct(species, .keep_all = TRUE) %>%
  # rename transfers to destination status
  mutate(status = ifelse(grepl("to Extinct",status),"Extinct",
                   ifelse(grepl("to Critically Endangered",status),"Critically Endangered",
                   ifelse(grepl("to Endangered",status),"Endangered",
                   ifelse(grepl("to Vulnerable",status),"Vulnerable", status))))) %>%
  # remove species delisted
  filter(!species %in% remove)

# Flora 1472 + Fauna 666 = 2138 listed species
nrow(epbc_2024)

# append to 2000-2023 lists
epbc_updated <- rbind(epbc_all, epbc_2024) %>%
  arrange(species)

# taxonomic changes in 2024 (I manually keep these as I find them)
taxon <- read_csv("data/epbc_taxonomy.csv") %>%
  clean_names() %>%
  # drop other names
  select(!c(synonymous_listing, added)) %>%
  filter(!accepted_name == "")

epbc_updated <- epbc_updated %>%
  left_join(taxon, by = c("species" = "listed_as")) %>%
  mutate(
    species = case_when(
      !is.na(accepted_name) ~ accepted_name,  # Use new name if available
      TRUE ~ species                # Keep original if no update needed
    )
  ) %>%
  select(-accepted_name) %>%
  # remove duplicate records
  distinct(species, year, .keep_all = TRUE)
  
# write out
write.csv(epbc_updated, "data/epbc_processed_2024.csv", row.names = FALSE)
```

## Summary

I summarise the number of species per threat category per year. For graphing simplicity we lump "extinct in the wild" with "extinct" and "conservation dependent" with "vulnerable".

```{r}
# read in processed data to this calendar year
epbc <- read_csv("data/epbc_processed_2024.csv") %>%
  # lump categories
  mutate(status = ifelse(grepl("Extinct in the wild",status),"Extinct",
                         ifelse(grepl("Conservation Dependent",status),"Vulnerable",status))) %>%
  group_by(year, status) %>%
  summarise(count = length(species))
```

I make the bar chart in excel for ease of formatting by the graphic design team for the report. But I'll make one here too. 

```{r}
# reorder epbc for plot
epbc <- epbc %>%
  mutate(status = factor(status, levels = c("Extinct", "Critically Endangered", "Endangered", "Vulnerable")))

# stacked bar chart
ggplot(epbc)+
  geom_bar(aes(x = year, y = count, group = status, fill = status),
           position="stack", stat="identity")+
  theme_classic()+
  xlab(element_blank())+
  ylab("Count of species")+
  scale_fill_manual(values = c("gray5", "firebrick4", "tan1", "lightgoldenrod"))
```

Wide format for excel.

```{r}
# pivot wider
epbc_export <- epbc %>%
  pivot_wider(names_from = status, values_from = count) %>%
  # add total column
  mutate(total = rowSums(across(where(is.numeric))))

# write out
write.csv(epbc_export, "summary/epbc_summary_2024.csv", row.names = FALSE)
```

## IBRA subregion map

In addition to the time series stacked bar chart, we make an annual map of count of threatened species by [biographic subregion](https://fed.dcceew.gov.au/datasets/14495298f2744efd8960ff952a15ee9d).We get the spatial distribution of threatened species from the [SNES (Species of National Environmental Significance) public database - last version updated 23/09/2024](https://fed.dcceew.gov.au/datasets/9d313bb078b9421ebebc835b3a69c470).

```{r}
# ibra subregions
ibra <- vect("shapefiles/ibra61_reg/ibra61_reg.shp") %>%
  clean_names() %>%
  select(reg_name)

# SNES geodatabase - only on local, not on github repo
snes <- vect("shapefiles/snes_public_gdb.gdb") %>%
  clean_names() %>%
  # keep only threatened species listings
  filter(threatened_status %in% c("Critically Endangered", "Conservation Dependent",
                           "Vulnerable", "Endangered")) %>%
  # match IBRA projection
  project(crs(ibra))

# and SNES as a dataframe to reduce computation when geometry not needed
snes_df <- snes %>%
  st_as_sf() %>%
  st_drop_geometry()

length(unique(snes_df$scientific_name))
```

SNES dataframe has 2034 species, the 2024 list of extant species is 2036 - so which missing?  

```{r}
# read in epbc taxonomy to standardise names
taxon <- read_csv("data/epbc_taxonomy.csv") %>%
  clean_names() %>%
  # drop other names
  select(!synonymous_listing) %>%
  filter(!accepted_name == "")

# read in current epbc list
epbc <- read_csv("data/epbc_processed_2024.csv") %>%
  # filter to 2024 and extant
  filter(year == 2024 & !status == "Extinct") %>%
  # keep just names
  select(species, name)

# snes sci names list - standardised to epbc list
snes_names <- snes_df %>%
  select(scientific_name, vernacular_name) %>%
  rename(name = vernacular_name) %>%
  distinct() %>%
  left_join(taxon, by = c("scientific_name" = "listed_as")) %>%
  mutate(
    # Remove subspecies notation
    scientific_name = str_remove(scientific_name, "subsp\\. "),
    # Create standardised name
    species = case_when(
      # If there's an accepted name in taxon, use that
      !is.na(accepted_name) ~ accepted_name,
      # If the name is in epbc list, keep it as is (with parentheses)
      scientific_name %in% epbc$species ~ scientific_name,
      # Otherwise remove parenthetical text
      TRUE ~ str_remove_all(scientific_name, "\\s*\\([^)]+\\)")
    )
  ) %>%
  # drop old names
  select(!c(scientific_name, accepted_name))

# Find mismatches in both directions
mismatches <- bind_rows(
  snes_names %>% 
    select(name = species) %>%
    mutate(source = "snes_only") %>%
    anti_join(epbc %>% select(name = species)),
  epbc %>% 
    select(name = species) %>%
    mutate(source = "epbc_only") %>%
    anti_join(snes_names %>% select(name = species))
) %>%
  arrange(name)
```

The SNES polygons are missing Galaxias pedderensis and Lomatia tasmanica. Ah, that's because Galaxias pedderensis is extinct in the wild - so no polygon to be in. For the Lomatia, the SPRAT listing says "The entire population is contained within the Southwest National Park that is part of the Tasmanian Wilderness World Heritage Area" so we can manually add a record to the matching IBRA subregion, which is Tasmanian West (TWE).

For the map, first we count terrestrial species by IBRA subregion.

```{r}
# filter to terrestrial i.e. not listed as marine or overfly marine
snes_terra <- snes %>%
  filter(is.na(marine) | marine == "Listed - overfly marine area")

# find which snes polygons intersect with which ibra regions
ibra_ts <- terra::intersect(snes_terra, ibra)

# count of terrestrial ts by ibra subregion intersects
ibra_ts_count <- ibra_ts %>%
  st_as_sf() %>%
  st_drop_geometry() %>%
  select(scientific_name, reg_name) %>%
  distinct() %>%
  group_by(reg_name) %>%
  summarise(ts_count = length(scientific_name)) %>%
  # add one to Tasmanian West for Lomatia tasmanica
  mutate(ts_count = ifelse(reg_name=="Tasmanian West", ts_count + 1, ts_count))

# polygons with counts
ibra_ts_count_map <- left_join(ibra, ibra_ts_count)
```

Then we count species from Australian waters and island territories, to display as point on map background.

```{r}
# count marine species
sp_marine <- snes_df %>%
  filter(marine == "Listed") %>%
  select(scientific_name) %>%
  distinct()

nrow(sp_marine)

# count species by external territories
juris <- data.frame(acronym = c("ACI", "CKI", "CI", "CSI", "NFI", "HMI", "AAT"),
                    territory = c("Ashmore and Cartier Islands", "Cocos Keeling Islands", 
                                  "Christmas Island", "Coral Sea Islands", "Norfolk Island",
                                  "Heard and Macdonald Islands", "Australian Antarctic Territory"))

ts_juris <- data.frame()

for(i in 1:7){
   sp <- snes_df %>%
    filter(grepl(juris$acronym[i], regions)) %>%
    distinct(scientific_name)
   
   temp <- data.frame(name = juris$territory[i],
                          count = nrow(sp))
   
   ts_juris <- rbind(ts_juris, temp)
}

# combine external territories and Marine
territories <- ts_juris %>%
  rbind(data.frame(name = "Australian waters", count = nrow(sp_marine)))
```

Finally make the map, with terrestrial threatened species by IBRA subregion and inset points for external territories and marine listings. 

```{r}
# using the package showtext to get the custom font
font_add(family = "UniNeue-Trial-Book", regular = "fonts/UniNeue-Trial-Book.ttf")
showtext_auto()

# ggplot ibra subregion with species count, and points and labels for external territories
map_with_territories <- 
  ggplot() +
  geom_spatvector(data = ibra_ts_count_map, aes(fill = ts_count), colour = NA) +
  scale_fill_viridis_c(
    option = "inferno", 
    direction = -1,
    guide = guide_colorbar(
      title = NULL,
      label.theme = element_text(size = 24, family = "UniNeue-Trial-Book")
    )) +
  theme_void() +
  
  geom_point(
    data = territories,
    aes(x = 154, y = -45 + (seq_along(name) * 1.5),
        fill = count),
    colour = "white",
    size = 5,
    shape = 21,
    show.legend = FALSE
  ) +
  geom_text(
    data = territories,
    aes(x = 155, y = -45 + (seq_along(name) * 1.5)),
    label = territories$name,
    hjust = 0,
    vjust = 0.5,
    size = 7,  # Increased from 3.5 to 5
    family = "UniNeue-Trial-Book"  # Make sure to specify font family here
  ) +
  coord_sf(clip = "off") +
  theme(plot.margin = margin(r = 150, b = 20))

# save
ggsave("map_with_territories.png", 
       map_with_territories, 
       width = 8,    # Increased width for better text spacing
       height = 5,  # Slightly increased height
       dpi = 400)
```

# Coral cover

In the oceans section of the report we include annual trends in Great Barrier Reef coral cover. Coral cover is collected by the Australian Institute of Marine Science (AIMS) using a benthos manta tow method.

I downloaded the publicly available data from the [AIMS data repository](https://apps.aims.gov.au/metadata/view/5bb9a340-4ade-11dc-8f56-00008a07204e), which is currently for 1993-2023. 

Australian Institute of Marine Science (AIMS). (2015). AIMS Long-term Monitoring Program: Crown-of-thorns starfish and benthos Manta Tow Data (Great Barrier Reef). https://doi.org/10.25845/5c09b0abf315a, accessed 15-Feb-2025.

```{r}
coral_past <- read_csv("data/manta-tow-1993-2023.csv") %>%
  clean_names() %>%
  group_by(report_year) %>%
  summarise(coral_cover = mean(mean_live_coral),
            sd = sd(mean_live_coral),
            reef_count = length(report_year),
            ci = as.numeric(1.96*(sd/sqrt(reef_count))))
```

In the future I can just append the annual results to the previous processed output. 

To get the 2024 values I manually imported reports from the [reef data viewer](https://apps.aims.gov.au/reef-monitoring/sector/list) by sector and copying the 2024 table.

```{r}
# manually downloaded 2024 data
coral_2024 <- read_csv("data/manta-tow-2024.csv") %>%
  clean_names() %>%
  group_by(report_year) %>%
  summarise(coral_cover = mean(live_hard_coral_cover_percent),
            sd = sd(live_hard_coral_cover_percent),
            reef_count = length(report_year),
            ci = as.numeric(1.96*(sd/sqrt(reef_count))))

# add 2023 to previous data
coral <- rbind(coral_past, coral_2024)

# write out 
write.csv(coral, "data/coral_processed_2024.csv", row.names = FALSE)
```

And plot (again, this will be done by the design team but I want to mock up here because I'm curious).

```{r}
ggplot(data=coral)+
  geom_ribbon(aes(ymin = coral_cover - ci, 
                  ymax = coral_cover + ci, 
                  x = report_year), fill = "skyblue", alpha = .4)+
  geom_point(data = subset(coral, report_year==2024), aes(report_year, coral_cover))+
  geom_line(aes(report_year, coral_cover))+
  theme_classic()+
  ylim(0, 50)
```

# Cane toad spread

For the 2023 report we included a special figure on cane toad spread using [Atlas of Living Australia (ALA) data](https://www.ala.org.au/).

First, configure galah credentials to access ALA data. 

```{r}
alah_config(email = readChar("apikey_google.txt", nchars = file.info("apikey_google.txt")$size))
```

Next, call a download of cane toad records applying filtering.

```{r}
# call data
toads_ala <- galah_call() %>%
  galah_identify("Rhinella marina") %>%
  galah_select(group = "basic",
               basisOfRecord,
               coordinateUncertaintyInMeters,
               occurrenceStatus) %>%
  atlas_occurrences() %>%
  clean_names()

# save ala data so I dont need to repeat the download call
write.csv(toads_ala, "Rhinella_marina_ala.csv", row.names = FALSE)

# apply additional filtering 
toads <- read_csv("Rhinella_marina_ala.csv") %>%
  filter(occurrence_status == "PRESENT", # remove absences
         basis_of_record %in% c("OCCURRENCE", "OBSERVATION", #obs only
                                "HUMAN_OBSERVATION","MACHINE_OBSERVATION"),
         !is.na(event_date), #remove no date
         !is.na(decimal_longitude),
         !is.na(decimal_latitude),
         !str_ends(1, pattern = "000"), # remove dubious coords
         !str_ends(2, pattern = "000"),
         !coordinate_uncertainty_in_meters>2000) %>% # remove >2km uncertainty
  mutate(year = year(event_date))
```

Next rasterise convex hulls by time windows.

```{r}
# rasterise by time bin, convert to concave hull and smooth range map
base_raster <- rast(ncol=62.5, nrow=52.3, xmin=105.5, xmax=168, ymin=-43, ymax=-9.3)
ch_ratio <- c(.25, .22, .15, .11)
yearA <- c(1970, 2000, 2020, 2024)
yearB <- c(1900, 1970, 2000, 2020)
toad_maps <- data.frame()

for(i in 1:4){
  temp <- subset(toads, year<yearA[i] & year>=yearB[i]) %>%
    select(c(5,4)) %>%
    data.matrix() %>%
    rasterize(base_raster, fun = "max") %>%
    as.polygons() %>%
    st_as_sf() %>%
    drop_crumbs(threshold = units::set_units(10000, km^2)) %>% # drop single cell polygons
    st_concave_hull(ratio = ch_ratio[i]) %>% # variable concave hull ratio by time period
    smooth(method = "ksmooth") %>%
    vect() %>%
    crop(oz) %>%
    st_as_sf()  %>%
    mutate(bin = paste("pre", yearA[i], sep = "_"),
           factor = i ) %>%
    select(!max)
  
  toad_maps <- rbind(toad_maps, temp)
}

toad_maps <- arrange(toad_maps, -factor)
```

Finally plot map.

```{r}
ggplot()+
  geom_sf(data = toad_maps, aes(fill = bin), colour = NA)+
  geom_sf(data = oz, fill = NA)+
  theme_void()+
  theme(legend.position = c(.85, .5))+
  #geom_point(data = toads, aes(decimal_longitude, decimal_latitude))
  scale_fill_manual(values = c("wheat", "tan2", "darkorange3","brown"),
                     labels = c("Pre 1970", "1970-1999",
                                "2000-2019", "2020-2023"))+
  labs(fill = "Time Period")
```

